Parsing Postfix log files

### 2007-06-04 doc/logparser.tex
logparser.tex: * Lots of small corrections and improvements throughout the text. * Lots of place-holder sections. * get the bibliography properly into the TOC.
logparser-flow-chart.dot: changes in wording.
logparser-bibliography.bib: added "Last checked" as a note to all entries.
Makefile: stop the make if latex or bibtex fail.

% Better typesetting of URLs.
\usepackage{url}
% Include images
\usepackage[final]{graphicx}
% Change how nested enumerate environments are labelled.
\renewcommand{\labelenumii}{\roman{enumii}:}

@@

XXX FIGURE OUT WHEN TO USE MAIL AND WHEN TO USE CONNECTION\@; POSSIBLY STATE
THAT THEY'RE MORE OR LESS EQUIVALENT

@@

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug a problem,
explain to a user why their mail is being rejected, or check whether their
new anti-spam measures are working.  The more adventurous will have
generated statistics to show how many hits each of their anti-spam measures
has gotten in the last week, and possibly even generated some graphs to
clearly illustrate the point to management or complaining
users.\footnote{This was the author's first real foray into processing
Postfix logs.}  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing lines independently.
One of the barriers to this type of processing is the unstructured nature
of Postfix logs, where each logging line was added on an ad hoc basis as a
requirement was discovered or new functionality was added.  Further
complication because the list of rejection messages is not fixed: new
messages can be added by the administrator; every Real-time Black List
returns a different explanatory message; policy servers may have different
messages depending on the characteristics of the connection; there are many
ways in which the log lines may differ.  This paper documents the difficult
process of parsing Postfix logs, presenting a program which parses logs and
places the resulting data into a database.  The gathered data can then be
used to optimise current anti-spam measures, provide a baseline to test new
anti-spam measures against, or to produce statistics showing how effective
those measures are.  There are numerous other uses for such data: improving
server performance by identifying troublesome destinations and
reconfiguring appropriately or dedicating transports to those destinations;
identifying high volume uses (e.g.\ customer newsletters) and restricting
those uses to off-peak times; as a base for billing customers on a shared
server; etc.

Section~\ref{background} provides background information about the
following: the motivation for this project; the Postfix Mail Transport
Agent; a short overview of the parser; the assumptions inherent in the
design of the algorithm and program; using a database schema as an API\@;
and a comparison against other Postfix log parsers.

Section~\ref{rules} discusses the parsing rules in detail, explaining their
structure, giving an example rule and sample data it matches successfully
against.

Section~\ref{parsing-algorithm} contains the meat of the paper, describing
a naive parsing algorithm and the complications encountered which shaped
the full algorithm, followed by a comprehensive explanation of the
different stages of the algorithm and the actions taken during the
execution of the algorithm.

@@

Section~\ref{references} catalogues resources used in developing the
algorithm, writing the program and preparing this paper.  
Any resources referred to in the paper are also given, plus some additional
resources expected to be helpful in understanding Postfix, anti-spam
techniques, or the paper.

@@

A short example of the optimisation possible using data from the database
is optimal ordering of Postfix restrictions:

@@

    \item Libraries exist for the majority of programming languages
        allowing access to databases, whereas if this parser was written
        as a library a new interface layer would need to be written for
        every programming language using API\@.

    \item Databases provide complex querying and sorting functionality to
        the user without requiring large amounts of programming.  All
        databases provide one or more programs, of variable complexity and
        sophistication, which can be used for ad hoc queries with minimal
        investment of time.

@@

@@

@@

Postfix is a highly configurable, high performance, secure and scalable
Mail Transport Agent.  It features extensive optional anti-spam
restrictions, allowing an administrator to deploy those restrictions which
they judge suitable for their site's needs, rather than a fixed set chosen
by Postfix's author.  These restrictions can be selectively applied,
combined and bypassed on a per-client, per-recipient or per-sender basis,
allowing varying levels of defense and/or permissiveness.  Postfix
leverages simple lookup tables to support arbitrarily complicated
user-defined sequences of restrictions and exceptions, with policy
servers\footnote{Policy servers will be explained shortly.} as the ultimate
in flexibility.  Administrators can also supply their own rejection
messages to make it clear to senders why exactly their mail was rejected.
Unfortunately this flexibility has a cost: complexity in the logs
generated.  While it is easy to use \texttt{grep(1)} to determine the fate
of an individual email, following the journey an email takes through
Postfix can be quite difficult.  The logs tend to follow a 90\%-10\%
pattern: 90\% of the time the journey is simple, but the other 10\% of the
time requires 90\% of the code.\footnote{These numbers don't have a solid
scientific basis, they're based on gut feeling from writing and debugging
the software.}

@@

A policy server \cite{policy-servers} is an external program that accepts
state information from Postfix for each SMTP command not rejected by an
earlier restriction; the policy server can utilise that state information
to implement whatever logic is required.  E.g.\ some users can be
restricted to sending mail on the third Tuesday after pay day only --- this
example may actually be useful in a payroll system to prevent spam or
(worse) phishing mails with faked sender addresses.  More commonly
encountered scenarios are:

@@

    \item Checking Sender Policy Framework records \cite{openspf},
        \cite{wikipedia-spf}.  SPF records specify which mail servers are
        allowed to send mail claiming to be from a particular domain.  The
        intention is to reduce spam from faked sender addresses,
        backscatter \cite{postfix-backscatter} and joe-jobs
        \cite{wikipedia-joe-job}; however there
        has been a lot of resistance to the proposal because it breaks or
        vastly complicates some features of SMTP.

    \item Greylisting \cite{greylisting}, a technique that temporarily
        rejects mail where the triple of (sender, recipient, remote IP
        address) is unknown; the assumption is that it's uneconomical for a
        spammer to retry, but that a legitimate mail server must retry.
        Sadly spammers are using increasingly complex and well written
        programs to distribute spam, frequently using an ISP provided SMTP
        server from a compromised machine on its network.  Greylisting is
        slowly becoming less useful, but it does block a large percentage
        of spam mail at the moment.

    \item Using a scoring system such as Policyd-weight
        \cite{policyd-weight} where tests accumulate points against the
        sending system --- if the eventual score is too high the mail is
        rejected.

    \item Rate limiting or throttling on a per-sender, per-client or
        per-recipient basis as performed by Policyd \cite{policyd}.

@@

Example attributes taken from \cite{policy-servers}:

@@

Before getting into detail about the parser a brief overview is in order.
The parser is split into two parts: the parsing algorithm and the rules
which are applied to the lines.  Rules can be thought of as the Lex part of
a Lex and YACC style parser; rules identify the line and return some data,
which the algorithm (the YACC part) receives and performs the requested
action.  Rules are solely concerned with identifying a line and extracting
data from it, whereas the algorithm's task is to follow the journey each
mail takes through Postfix, piecing the data together into a coherent
whole, saving it in a useful and consistent form, and performing
housekeeping duties.  Readers familiar with context-free grammars may find
Section \ref{comparison against context-free grammars} helpful.

@@

    \item The logs are whole and complete; nothing has been removed, either
        deliberately or accidentally (e.g.\ log rotation gone awry, file
        system filling up, logging system unable to cope with the volume of
        logs).

@@

10 other parsers have been reviewed in Appendix~\ref{other-parsers} as part
of the background research for this project.  None of the reviewed parsers
perform the type of advanced parsing and log correlation described here;
all are intended to perform a specific parsing and reporting task, rather
than be a generic parser and leave generating reports from the data to
other programs.  Some save the data extracted to a database but the
majority discard all data once finished running, making historical analysis
impossible.  The other parsers listed all produce a report of greater or
lesser complexity and detail, whereas the program described here doesn't
attempt to produce a report at all; that responsibility is deferred to a
separate program, to be developed later.  The parsing algorithm and program
described here are designed to enable much more detailed log analysis by
providing a stable platform for subsequent programs to develop upon.

@@

The complexity and variation in Postfix's logs requires similar flexibility
in the parser; decoupling the parsing rules from the associated actions
allows new rules to be written and tested without requiring modifications
to the algorithm source code (significantly lowering the barrier to entry
for new or casual users), and greatly simplifies both algorithm and rules.
It also creates a clear separation of functionality: rules handle low level
details of identifying and extracting data from a line, whereas the
algorithm handles the higher level details of following the path a mail
takes through Postfix, assembling the required data, etc.

@@

XXX DESCRIBE IT IN LEFT: RIGHT PRODUCTION TERMS, IF POSSIBLE.

@@

    \item [description] Something must have occurred to cause Postfix to
        log each line (e.g. a remote client connecting causes a connection
        line to be logged).  The description field is generally used to
        describe the action causing the log lines this rule matches.

@@

@@

        For efficiency the keywords are expanded and every rule's regex is
        compiled before attempting to parse the log file --- otherwise each
        regex would be recompiled each time it was used, resulting in a
        large, data dependent slowdown.  Graph \ref{normal regex vs discard
        regex} shows execution times gathered over 93 log files, with and
        without caching the regex; also shown on the graph are the file
        sizes and number of lines in each file.  Graph \ref{normal regex vs
        disarded regex factor} shows the percentage execution time increase
        when not caching each regex.  Each log file was parsed 10 times,
        the first run discarded and the remaining 9 runs averaged.  Logging
        to the database was disabled for the test runs.

@@

        \texttt{client\_hostname, client\_ip, server\_ip, server\_hostname,
        \newline helo, sender, recipient, smtp\_code, data and child.}

    \item [result\_data, connection\_data] Sometimes rules need to supply a
        piece of data which isn't present in the log line: e.g.\ setting
        \texttt{smtp\_code} when mail is accepted.  The format and allowed
        variables are the same as for \texttt{result\_cols} and
        \texttt{connection\_cols}, except that arbitrary
        data\footnote{Commas and semi-colons cannot be escaped and thus
        cannot be used.  This is intended for small amounts of data rather
        than large, so dealing with escape sequences seemed unnecessary.}
        is permitted on the right hand side of the assignment.

            \item [info] Represents an unspecified intermediate action that
                the parser is not interested in per se, but which does log
                useful information, supplementing other log lines.

            \item [ignored] An action which is not only uninteresting in
                itself, but which also provides no useful data.

    \item [action] The action the algorithm will take, e.g.
        \begin{description}

            \item [ignore] Ignore this line.  The simplest action, it is
                required because every line from a program of interest to
                the parser must be parsed properly; any unparsed line is
                considered an error.

            \item [commit] The mail is finished with, enter it in the
                database, and remove the mail from memory.

        A full list can be found in Section~\ref{actions-in-detail}.

    \item [queueid] Specifies the match from the regex which gives the
        queueid, or zero if the log line doesn't contain a queueid.

    \item [rule\_order] This is an efficiency measure.  This counter is
        maintained for every rule and incremented each time the rule
        successfully matches.  At the start of each run the program sorts
        the rules in descending rule\_order, and at the end of the run
        updates every rule's rule\_order.  Assuming that the distribution
        of log lines is reasonably consistent between log files, rules
        matching more commonly occurring log lines will be tried before
        rules matching less commonly occurring log lines, lowering the
        program's execution time.  Graph \ref{normal vs shuffled vs
        reversed ordering} plots the execution time over 93 log files for
        normal, shuffle and reverse ordering.  Graph \ref{normal vs
        shuffled vs reversed ordering factor} shows the percentage
        execution time increase for shuffled and reversed over normal.
        Each log file was parsed 10 times, the first run discarded and the
        remaining 9 runs averaged.  Logging to the database was disabled
        for the test runs.

        It is hoped that this will prove to have greater effectiveness as
        the number of rules grows; currently the effect is negligible
        in comparison to caching each compiled regex, but it still provides
        approximately a 7\% -- 10\% performance increase.

    \item [priority] This is the user-configurable companion to
        rule\_order: rules with a higher priority will be tried first,
        overriding rule\_order.  This allows more specific rules to take
        precedence over more general rules.

@@

This example rule matches the message Postfix logs when it rejects mail
from a sender address where the domain has neither an MX record nor an A
record, i.e.\ mail could not be delivered to the sender's address.  For full
details see
http://www.postfix.org/postconf.5.html\#reject\_unknown\_sender\_domain

@@

    \item [name, description, restriction\_name and postfix\_action:] are
        not \newline used by the algorithm, they serve to document the rule.

    \item [program and regex:] If the program equals the program which
        logged the line the regex will be matched against the line; if
        successful the action will be called, if not the next rule will be
        tried.

    \item [action:] will be called if the regex matches successfully (see
        Section \ref{actions-in-detail} for full details).

@@

    \item [queueid:] is used the extract the queueid matched by the regex,
        so that mails can be found by queueid and actions performed on the
        mail.

@@

@@

Unfortunately it's not that easy.

@@

    \item The mail lacks a queueid until it has been accepted, so log lines
        must first be correlated by the smtpd pid, then transition to being
        correlated by the queueid.  This is relatively minor, but does
        require:

@@

            \item Two versions of several functions, \texttt{by\_pid} and
                \texttt{by\_queueid}.

            \item Two containing data structures to hold the data
                structure for each connection.

            \item Most importantly: every section of code must know whether
                it is needs to lookup the data structures by pid or
                queueid.

@@

    \item Multiple independent mails may be delivered during one
        connection: this requires cloning the current data as soon as a
        mail is accepted, so that subsequent mails won't trample over each
        other's data.  This must be done every time a mail is accepted, as
        it's impossible to tell in advance which connections will accept
        multiple mails.  It is quite easy to overlook this complication
        because only a small minority of connections accept more than one
        mail. Happily once the mail has been accepted log entries won't be
        correlated by pid for that mail any more (its queueid will be used
        instead), so there isn't any ambiguity about which mail a given log
        line belongs to.  The original connection will be discarded unsaved
        when the client disconnects.  One unsolved difficulty is
        distinguishing between different groups of rejections, e.g.\ when
        dealing with the following sequence:

@@

    \item The most difficult complication is that mails are not always
        delivered directly to a mailbox (or program; there is very little
        difference between the two): sometimes they are accepted for a
        local address but need to be delivered to one or more remote
        addresses due to aliases.  When this occurs a child mail will be
        injected into the postfix queue, but without the explicit logging
        smtpd or sendmail injected mails have, so the source is not
        immediately discernible from the log line in which the mail first
        appears; from a strictly linear reading of the logs it
        \textit{usually\/} appears as if the child mail has appeared from
        thin air.  Subsequently the parent mail will log the creation of
        the child mail:

@@

        Because of this the parser cannot complain when it encounters a log
        line from qmgr for a previously unseen mail; it must flag the mail
        as coming from an unknown origin and subsequently clear the flag if and
        when the origin of the mail becomes clear.  Obviously the parser
        could omit checking of where mails originate from, but I feel that
        it is better to require an explicit source, as bugs are more likely
        to be exposed.

        Process scheduling can have a still more confusing effect: quite
        often the child mail will be created, delivered and entirely
        finished with \textbf{before} the parent logs the creation line!
        Thus mails flagged as coming from an unknown origin cannot be
        entered into the database when their final log line is parsed;
        instead they must be marked as database ready and subsequently
        entered by the parent mail once it has been identified.  Quite
        apart from identifying mail injected in an unknown fashion, or bugs
        in the parser, unknown origin mails need to be marked as such
        because they lack some data present in the parent mail; they must
        copy that data from their parent before being entered in the
        database.

        The algorithm requirements are discussed in
        section~\ref{tracking-re-injected-mail}.

@@

The intermingling of log entries from different mails immediately rules out
the possibility of handling each mail in isolation; the parser must be
capable of handling multiple mails in parallel, each potentially at a
different stage in its journey, without any interference between mails ---
except in the minority of cases where intra-mail interference is required.
The best way to implement this is to maintain state information for every
unfinished mail and manipulate the appropriate mail correctly for each log
line encountered.  The parser thus requires both a method of mapping log
lines to the correct mail and a method of specifying the action the log
line represents.  The former is achieved by using the smtpd process id to
identify the correct mail during the initial phase, then switching to the
queueid once the mail has been accepted.  The latter uses the action field
of the rule which matched the log line, executing the code in the function
named by the action.  From a high level viewpoint this design shares
significant similarity with how an editor maintains state for multiple
files and performs the actions the user initiates on the correct file.

Section~\ref{actions-in-detail} explains the actions in substantive detail;
this section will omit such detail because it would clutter and confuse the
algorithm description.  The data flow chart in section~\ref{flow-chart}
should be consulted while reading this section also.

@@

Everything starts off with a mail entering the system, whether by local
submission via postdrop/sendmail, by SMTP, or by re-injection due to
forwarding.  Local submission is the simplest of the three: a queueid is
assigned immediately and the sender address is logged (action: pickup;
flowchart:~2).

@@

    \item This is followed by rejection of sender, recipients, client
        address, etc. (action: save\_by\_pid; flowchart:~4); acceptance of
        one or more mails (action: clone, then save\_by\_pid;
        flowchart:~5,~4); or some interleaving of both.

    \item The client disconnects (action: disconnect; flowchart:~6).  As
        soon as a mail is accepted it is assigned a queueid; thus if there
        is no queueid Postfix didn't accept any mail over that connection,
        it rejected all attempted deliveries due to the configured
        restrictions.  Absence of a queueid is checked for during
        disconnection handling (flowchart:~7) and if none is found the data
        is cleaned up and entered in the database (flowchart:~8;
        subroutines: fixup\_connection and commit\_connection), then
        removed from memory (flowchart:~9).

    \item If one or more mails were accepted there will be more log entries
        later, see section~\ref{mail-delivery}.

@@

The algorithm for tracking and saving re-injected mail to the database will
finally be described:

@@

This list contains all the actions available in the algorithm.  Each action
is passed the same arguments:

@@

>>>>

% Better typesetting of URLs.
\usepackage{url}
% Include images
\usepackage[final]{graphicx}
% Change how nested enumerate environments are labelled.
\renewcommand{\labelenumii}{\roman{enumii}:}
\renewcommand{\refname}{Bibliography}
\usepackage[section,numbib,nottoc]{tocbibind}

@@

XXX LOOK AT citeseer.com AND scholar.google.com FOR RELATED PAPERS

@@

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to debug a problem,
explain to a user why their mail is being rejected, or check whether their
new anti-spam measures are working.  The more adventurous will have
generated statistics to show how many hits each of their anti-spam measures
has gotten in the last week, and possibly even generated some graphs to
clearly illustrate the point to management or complaining
users.\footnote{This was the author's first real foray into processing
Postfix logs.}  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing must correlate the log lines
per-connection or per-queueid rather than processing lines independently.
One of the barriers to this type of processing is the unstructured nature
of Postfix logs, where each logging line was added on an ad hoc basis as a
requirement was discovered or new functionality was added.  Further
complication arises because the list of rejection messages is not fixed:
new messages can be added by the administrator with custom checks; every
Real-time Black List returns a different explanatory message; policy
servers may have different messages depending on the characteristics of the
connection; there are many ways in which the log lines may differ between
servers.  This paper documents the difficult process of parsing Postfix
logs, presenting a program which parses logs and places the resulting data
into a database.  The gathered data can then be used to optimise current
anti-spam measures, provide a baseline to test new anti-spam measures
against, or to produce statistics showing how effective those measures are.
There are numerous other uses for such data: improving server performance
by identifying troublesome destinations and reconfiguring appropriately or
dedicating transports to those destinations; identifying regular high
volume uses (e.g.\ customer newsletters) and restricting those uses to
off-peak times; detecting virus outbreaks which propagate via email; as a
base for billing customers on a shared server.

\vspace{1em}\noindent\textbf{Layout of the paper:}

Section~\ref{background} provides background information useful in
understanding the paper, parser and algorithm.

Section~\ref{rules} discusses the parsing rules in detail, explaining their
structure, giving an example rule and sample data it matches successfully
against, followed by a discussion of rule efficiency concerns.

Section~\ref{parsing-algorithm} contains the meat of the paper, describing
a naive parsing algorithm and the complications encountered which shaped
the full algorithm, followed by a comprehensive explanation of the
different stages of the algorithm, the actions taken during the execution
of the algorithm, and further complications whose solution completes the
algorithm.

@@

Section~\ref{other-parsers} discusses other parsers and why they were
deemed unsuitable for the task.

Section~\ref{bibliography} is the bibliography, containing the resources
used in developing the algorithm, writing the program and preparing this
paper.  Also listed are some additional resources expected to be helpful in
understanding SMTP, Postfix, anti-spam techniques, or the paper.

Section~\ref{graphs} contains graphs illustrating the topics discussed in
section~\ref{rule efficiency}.

@@

A short example of the optimisation possible using data from the database
is which Postfix restrictions reject the highest number of mails:

@@

    \item Libraries exist for the majority of programming languages
        allowing access to databases, whereas if this parser was written
        as a library a new interface layer would need to be written for
        every programming language using the API\@.

    \item Databases provide complex querying and sorting functionality to
        the user without requiring large amounts of programming.  All
        databases provide a program, of variable complexity and
        sophistication, which can be used for ad hoc queries with minimal
        investment of time.

@@

    \item SQL is reasonably standard and many people will already be
        familiar with it; for those unfamiliar with it there are already
        lots of readily available resources from which to learn.  Although
        every vendor implements a different dialect of SQL, the basics are
        the same everywhere.

@@

\subsection{Database schema}
\label{database schema}

XXX WRITE THIS

\subsection{SMTP background}
\label{smtp background}

XXX WRITE THIS

@@

Postfix is a highly configurable, high performance, secure and scalable
Mail Transport Agent.  It features extensive optional anti-spam
restrictions, allowing an administrator to deploy those restrictions which
they judge suitable for their site's needs, rather than a fixed set chosen
by Postfix's author.  These restrictions can be selectively applied,
combined and bypassed on a per-client, per-recipient or per-sender basis,
allowing varying levels of severity and/or permissiveness.  Postfix
leverages simple lookup tables to support arbitrarily complicated
user-defined sequences of restrictions and exceptions, with policy
servers\footnote{Policy servers will be explained shortly.} as the ultimate
in flexibility.  Administrators can also supply their own rejection
messages to make it clear to senders why exactly their mail was rejected.
Unfortunately this flexibility has a cost: complexity in the logs
generated.  While it is easy to use \texttt{grep(1)} to determine the fate
of an individual email, following the journey an email takes through
Postfix can be quite difficult.  The logs tend to follow a 90\%-10\%
pattern: 90\% of the time the journey is simple, but the other 10\% of the
time requires 90\% of the code.\footnote{These numbers don't have a solid
scientific basis, they're based on gut feeling from writing and debugging
the software.}

\subsubsection{Mixing and matching Postfix restrictions}

XXX TO BE WRITTEN

@@

A policy server~\cite{policy-servers} is an external program that accepts
state information from Postfix for each SMTP command not rejected by an
earlier restriction; the policy server can utilise that state information
to implement whatever logic is required.  E.g.\ some users can be
restricted to sending mail on the third Tuesday after pay day only --- this
example may actually be useful in a payroll system to prevent problems from
spam or (worse) phishing mails with faked sender addresses.  More commonly
encountered scenarios are:

@@

    \item Checking Sender Policy Framework
        records~\cite{openspf},~\cite{wikipedia-spf}.  SPF records specify
        which mail servers are allowed to send mail claiming to be from a
        particular domain.  The intention is to reduce spam from faked
        sender addresses, backscatter~\cite{postfix-backscatter} and
        joe-jobs~\cite{wikipedia-joe-job}; however there has been a lot of
        resistance to the proposal because it breaks or vastly complicates
        some features of SMTP\@.

    \item Greylisting~\cite{greylisting}, is a technique that temporarily
        rejects mail when the triple of (sender, recipient, remote IP
        address) is unknown; on the second and subsequent delivery attempts
        from that triple the mail will be accepted.  The assumption is that
        retrying after a temporary failure is uneconomical for a spammer,
        but that a legitimate mail server must retry.  Sadly spammers are
        using increasingly complex and well written programs to distribute
        spam, frequently using an ISP provided SMTP server from a
        compromised machine on the ISP's network.  Greylisting is slowly
        becoming less useful, but it does block a large percentage of spam
        mail at the moment.

    \item Using a scoring system such as
        Policyd-weight~\cite{policyd-weight} where tests accumulate points
        against the sending system --- if the eventual score is too high
        the mail is rejected.

    \item Rate limiting or throttling on a per-sender, per-client or
        per-recipient basis as performed by Policyd~\cite{policyd}.

@@

Example attributes taken from~\cite{policy-servers}:

@@

Before getting into detail about the parser a brief overview is in order.
The parser is split into two parts: the parsing algorithm and the rules
which are applied to the lines.  Rules can be thought of as the Lex part of
a Lex and YACC style parser: rules identify the line and return some data,
which the algorithm (the YACC part) receives and uses when performing the
requested action.  Rules are solely concerned with identifying a line and
extracting data from it, whereas the algorithm's task is to follow the
journey each mail takes through Postfix, piecing the data together into a
coherent whole, saving it in a useful and consistent form, and performing
housekeeping duties.  Readers familiar with context-free grammars may find
Section~\ref{comparison against context-free grammars} helpful.

@@

    \item The logs are whole and complete: nothing has been removed, either
        deliberately or accidentally (e.g.\ log rotation gone awry, file
        system filling up, logging system unable to cope with the volume of
        logs).

@@

10 other parsers have been reviewed in Appendix~\ref{other-parsers} as part
of the background research for this project.  None of the reviewed parsers
perform the type of advanced parsing and log correlation described here;
all are intended to perform a specific parsing and reporting task, rather
than be a generic parser, extracting data and leaving generation of reports
from the data to other programs.  Some parsers save the data extracted to a
database but the majority discard all data once they have finished running,
making historical analysis impossible.  The other parsers reviewed all
produce a report of greater or lesser complexity and detail, whereas the
program described here doesn't attempt to produce a report at all; that
responsibility is deferred to a separate program, to be developed later.
The parsing algorithm and program described here are designed to enable
much more detailed log analysis by providing a stable platform for
subsequent programs to develop upon.

\subsection{Conventions used in the paper}

The words \textit{connection\/} and \textit{mail\/} are often used
interchangeably in this paper; in general the word used was chosen based on
the context it appears in.

@@

The complexity and variation in Postfix's logs requires similar flexibility
in the parser; decoupling the parsing rules from the associated actions
allows new rules to be written and tested without requiring modifications
to the algorithm source code (significantly lowering the barrier to entry
for new or casual users), and greatly simplifies both algorithm and rules.
It also creates a clear separation of functionality: rules handle low level
details of identifying log lines and extracting data from a line, whereas
the algorithm handles the higher level details of following the path a mail
takes through Postfix, assembling the required data, etc.

@@

XXX DESCRIBE IT IN LEFT\@: RIGHT PRODUCTION TERMS, IF POSSIBLE\@.

@@

    \item [description] Something must have occurred to cause Postfix to
        log each line (e.g.\ a remote client connecting causes a connection
        line to be logged).  The description field is generally used to
        describe the action causing the log lines this rule matches.

@@

            \item [INFO] Represents an unspecified intermediate action that
                the parser is not interested in per se, but which does log
                useful information, supplementing other log lines.

            \item [IGNORED] An action which is not only uninteresting in
                itself, but which also provides no useful data.

@@

        Additional fields are captured by \_\_RESTRICTION\_START\_\_, so
        rules using it will start the fields in result\_cols,
        connection\_cols, etc.\ at 5.

        For efficiency the keywords are expanded and every rule's regex is
        compiled before attempting to parse the log file --- otherwise each
        regex would be recompiled each time it was used, resulting in a
        large, data dependent slowdown.  Rule efficiency concerns are
        discussed in section~\ref{rule efficiency}.

@@

        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
        \newline server\_hostname and helo.\newline} \texttt{result\_cols:
        sender, recipient, smtp\_code, data \newline and child.}

    \item [result\_data, connection\_data] Sometimes rules need to supply a
        piece of data which isn't present in the log line: e.g.\ setting
        \texttt{smtp\_code} when mail is accepted.  The format and allowed
        variables are the same as for \texttt{result\_cols} and
        \texttt{connection\_cols}, except that arbitrary
        data\footnote{Commas and semi-colons cannot be escaped and thus
        cannot be used.  This is intended for use with small amounts of
        data rather than large, so dealing with escape sequences seemed
        unnecessary.} is permitted on the right hand side of the
        assignment.

    \item [action] The action the algorithm will take; a full list can be
        found in Section~\ref{actions-in-detail}.

    \item [queueid] Specifies the matching field from the regex which gives
        the queueid, or zero if the log line doesn't contain a queueid.

    \item [rule\_order] is an efficiency measure.  This counter is
        maintained for every rule and incremented each time the rule
        successfully matches.  At the start of each run the program sorts
        the rules in descending rule\_order, and at the end of the run
        updates every rule's rule\_order.  Assuming that the distribution
        of log lines is reasonably consistent between log files, rules
        matching more commonly occurring log lines will be tried before
        rules matching less commonly occurring log lines, lowering the
        program's execution time.  Rule efficiency concerns are discussed
        in section~\ref{rule efficiency}.

    \item [priority] This is the user-configurable companion to
        rule\_order: rules will be tried in order of priority, overriding
        rule\_order.  This allows more specific rules to take precedence
        over more general rules.

@@

This example rule matches the message logged by Postfix when it rejects
mail from a sender address where the domain has neither an MX record nor an
A record, i.e.\ mail could not be delivered to the sender's address --- for
full details see~\cite{reject-unknown-sender-domain}.

@@

    \item [name, description, restriction\_name and postfix\_action:] are
        not \newline used by the algorithm, they serve to document the rule
        for the user's benefit.

    \item [program and regex:] If the program in the rule equals the
        program which logged the line the regex will be matched against the
        line; if the match is successful the action will be called, if not
        the next rule will be tried.

    \item [action:] will be executed if the regex matches successfully (see
        section~\ref{actions-in-detail} for full details).

@@

    \item [queueid:] is used the extract the queueid matched by the regex,
        so that the mail can be found by queueid and actions performed on
        it.

@@

Additional fields are captured by \_\_RESTRICTION\_START\_\_, hence the
fields in result\_cols and connection\_cols start at 5.

\subsection{Rule efficiency}

\label{rule efficiency}

XXX WRITE ABOUT\@: DYNAMICS OF THE SYSTEM

XXX PRIORITY, OVERLAPPING RULES, MORE SPECIFIC VS \newline MORE GENERAL

XXX EXPLAIN THE DIPS IN THE GRAPHS WITH LARGE LOG FILES

XXX MENTION THAT THE PROGRAM HAS LINER COMPLEXITY AND SCALES LINEARLY WITH
THE LOG FILE SIZE

Graph~\ref{normal regex vs discard regex} shows execution times gathered
over 93 log files, with and without caching the regex; also shown on the
graph are the file sizes and number of lines in each file.
Graph~\ref{normal regex vs disarded regex factor} shows the percentage
execution time increase when not caching each regex.  Each log file was
parsed 10 times, the first run discarded and the remaining 9 runs averaged.
Logging to the database was disabled for the test runs.

Graph~\ref{normal vs shuffled vs reversed ordering} plots the execution
time over 93 log files for normal, shuffle and reverse ordering.
Graph~\ref{normal vs shuffled vs reversed ordering factor} shows the
percentage execution time increase for shuffled and reversed over normal.
Each log file was parsed 10 times, the first run discarded and the
remaining 9 runs averaged.  Logging to the database was disabled for the
test runs.

It is hoped that this will prove to have greater effectiveness as the
number of rules grows; currently the effect is negligible in comparison to
caching each compiled regex, but it still provides approximately a 7--10\%
performance increase.

@@

Unfortunately it's not that straightforward.

@@

\subsubsection{Queueid vs pid}

The mail lacks a queueid until it has been accepted, so log lines must
first be correlated by the smtpd pid, then transition to being correlated
by the queueid.  This is relatively minor, but does require:

@@

    \item Two versions of several functions: \texttt{by\_pid} and
        \texttt{by\_queueid}.

    \item Two state tables to hold the data structure for each connection.

    \item Most importantly: every section of code must know whether it
        needs to lookup the data structures by pid or queueid.

@@

\subsubsection{Connection reuse}

Multiple independent mails may be delivered during one connection: this
requires cloning the current data as soon as a mail is accepted, so that
subsequent mails won't trample over each other's data.  This must be done
every time a mail is accepted, as it's impossible to tell in advance which
connections will accept multiple mails.  It is quite easy to overlook this
complication because only a small minority of connections accept more than
one mail. Happily once the mail has been accepted log entries won't be
correlated by pid for that mail any more (its queueid will be used
instead), so there isn't any ambiguity about which mail a given log line
belongs to.\footnote{Unfortunately this statement is not completely
accurate: see section~\ref{timeouts-during-data-phase} for details.
However in general there isn't any ambiguity about which data structure
should be used for a given log line.}  The original connection will be
discarded unsaved when the client disconnects if there haven't been any
rejections.  One unsolved difficulty is
distinguishing between different groups of rejections, e.g.\ when dealing
with the following sequence:

@@

\subsubsection{Re-injected mails}

The most difficult complication is that locally addressed mails are not
always delivered directly to a mailbox: sometimes they are accepted for a
local address but need to be delivered to one or more remote addresses due
to aliases.  When this occurs a child mail will be injected into the
postfix queue, but without the explicit logging smtpd or sendmail injected
mails have.  Thus the source is not immediately discernible from the log
line in which the mail first appears; from a strictly chronological reading
of the logs it \textit{usually\/} appears as if the child mail has appeared
from thin air.  Subsequently the parent mail will log the creation of the
child mail:

@@

Because of this the parser cannot complain when it encounters a log line
from qmgr for a previously unseen mail; it must flag the mail as coming
from an unknown origin and subsequently clear the flag if and when the
origin of the mail becomes clear.  Obviously the parser could omit checking
of where mails originate from, but the author believes that it is better to
require an explicit source, as bugs are more likely to be exposed.

Process scheduling can have a still more confusing effect: quite often the
child mail will be created, delivered and entirely finished with
\textbf{before} the parent logs the creation line!  Thus mails flagged as
coming from an unknown origin cannot be entered into the database when
their final log line is parsed; instead they must be marked as database
ready and subsequently entered by the parent mail once it has been
identified.  Quite apart from identifying mail injected in an unknown
fashion, or bugs in the parser, unknown origin mails need to be marked as
such because they lack some data present in the parent mail; they must copy
that data from their parent before being entered in the database.
The effect of this complication upon the algorithm is discussed in
section~\ref{tracking-re-injected-mail}.

@@

The intermingling of log entries from different mails immediately rules out
the possibility of handling each mail in isolation; the parser must be
capable of handling multiple mails in parallel, each potentially at a
different stage in its journey, without any interference between mails ---
except in the minority of cases where intra-mail interference is required.
The best way to implement this is to maintain state information for every
unfinished mail and manipulate the appropriate mail correctly for each log
line encountered.  The parser thus requires both a method of mapping log
lines to the correct mail and a method of specifying the action the log
line represents.  The former is achieved by using the smtpd process id to
identify the correct mail during the initial phase, then switching to the
queueid once the mail has been accepted.  The latter uses the action field
of the rule which matched the log line, executing the code in the function
named by the action.  From a high level viewpoint this design shares
significant similarity with how an event dispatch loop works in a GUI\@.

Section~\ref{actions-in-detail} explains the actions in substantive detail;
this section will omit such detail because it would clutter and confuse the
algorithm description.  The data flow chart in section~\ref{flow-chart}
should also be consulted while reading this section.

@@

Everything starts off with a mail entering the system, whether by local
submission via postdrop/sendmail, by SMTP, by re-injection due to
forwarding, or internally generated by Postfix.  Local submission is the
simplest case: a queueid is assigned immediately and the sender address is
logged (action: pickup; flowchart:~2).

@@

    \item This is followed by rejection of sender, recipients, client
        address, etc. (action: rejection; flowchart:~4); acceptance of one
        or more mails (action: clone; flowchart:~5); or some interleaving
        of both.

    \item The client disconnects (action: disconnect; flowchart:~6).  If
        Postfix has rejected any SMTP commands the data will be saved to
        the database; if not there won't be any data to save (any mails
        accepted will already have been cloned so their data is in another
        data structure).

    \item If one or more mails were accepted there will be more log entries
        for those mails later, see section~\ref{mail-delivery}.

@@

The algorithm for tracking and saving re-injected mail to the database can
finally be described:

@@

Each action is passed the same arguments:

@@

XXX WRITE THIS
